---
title: "STRONG Study Primary Analyses"
author: "David A Hughes"
date: "2025-11-20"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

### load libraries

```{r, warning=FALSE, message=FALSE}
# install.packges("lmerTest")
# Bioconductor::install("lmerTest")
# BiocLite::install("lmerTest")

library(tidyverse)  # install.packages("tidyverse")
library(lmerTest)
library(readxl)
library(gtsummary)
library(gt)
library(emmeans)
library(dendextend)
library(grid)
```

### source function files

```{r}
files = list.files("functions", full.names = TRUE)
for(f in files){source(f)}
```


### load data

```{r}
f = "../data/STRONG_data spreadsheet.xlsx"
mydata = read_excel(f)

colnames(mydata) = gsub("\\/", "_", colnames(mydata) ) 
```

### Edit variable names

```{r}
colnames(mydata) = gsub("\\/", "_", colnames(mydata) ) 
colnames(mydata) = gsub(" ", "_", colnames(mydata) )
```

## Estiamte the number of effective outcome traits in the data

### Tree Cut

#### Correlation Structure Among all outcomes / variables

```{r, fig.width = 15, fig.height = 8}
vars_2_test = colnames(mydata)[-c(1:7)]
# length(vars_2_test) ## 39

Cmat = mydata |> select(vars_2_test) |> cor( use = "pairwise.complete.obs", method = "sp")
Dmat = as.dist( 1-abs(Cmat) )
nj = hclust(Dmat, method = "complete")

## k clusters based on cut height
k = cutree(nj, h = 0.5)
effnumb = length(unique(k)) ## 10


## Define Plot colors
pcol = c(RColorBrewer::brewer.pal(n = 4, name = "Set1"), "black")

## Make a dendrogram
nj <- suppressWarnings( as.dendrogram(nj) |>
  set("labels_cex", 1.25) |> 
  color_branches( h = 0.5, col = pcol) |>
  color_labels(h = 0.5, col = pcol) 
)

## report k
cli::cli_alert_info( "Number of effective outcome traits based on tree cutting: {effnumb}" )
```

```{r, include = FALSE}
## Save dendrogram to file
f = "../results/figures/variable_correlation_dendrogram.pdf"
pdf(f, width = 15, height = 8)
par(mar = c(17,3,2,1) )
plot(nj, 
     main = paste0("Clustering of Variables Based on Spearman's Correlation Structure; k = ",effnumb) )
abline(h = 0.5, col = "black", lwd = 2, lty = 2)
dev.off()
```


#### plot dendrogram

```{r, fig.width = 15, fig.height = 8}

## plot dendrogram
par(mar = c(17,3,2,1) )
plot(nj, 
     main = paste0("Clustering of Variables Based on Spearman's Correlation Structure; k = ",effnumb) )
abline(h = 0.5, col = "black", lwd = 2, lty = 2)

```
#### Correlation Structure Among non-meal variables

```{r, fig.width = 12, fig.height = 6, warning=FALSE, message=FALSE}
w = c( which(grepl("Breakfast", colnames(Cmat))),
       which(grepl("Lunch", colnames(Cmat))),
       which(grepl("Snack", colnames(Cmat))),
       which(grepl("Diner", colnames(Cmat)))
       # which(grepl("Sleep", colnames(Cmat)))
       )
Cmat2 = Cmat[-w,-w]
# dim(Cmat2)
Dmat2 = as.dist( 1-abs(Cmat2) ) 


nj = hclust(Dmat2, method = "complete")

## k clusters based on cut height
k2 = cutree(nj, h = 0.5)
effnumb2 = length(unique(k2))

pcol = c(RColorBrewer::brewer.pal(n = 4, name = "Set1"), "black")


## Make a dendrogram
nj <- as.dendrogram(nj) |>
  set("labels_cex", 1) |> 
  color_branches( h = 0.5, col = pcol) |>
  color_labels(h = 0.5, col = pcol)


## plot dendrogram
par(mar = c(13,3,2,1) )
plot(nj, 
     main = paste0("Clustering of Variables Based on Spearman's Correlation Structure; k = ", effnumb2 ) )
abline(h = 0.5, col = "black", lwd = 2, lty = 2)

```


### PCA

```{r}
pca_data = mydata |> select( all_of(vars_2_test) ) |> na.omit()
ind_pca_fit  = prcomp( pca_data, center = TRUE, scale = TRUE)
var_pca_fit  = prcomp( Cmat, center = TRUE, scale = TRUE)
## extract variance explained
varexp = summary(var_pca_fit)[[6]][2,]
## extract cummulative variance explained
cum_varexp = summary(var_pca_fit)[[6]][3,]
## Cummulative number of PCs needed to explain >=95% of the variance
effnumb = min( which( cum_varexp > 0.95 ) )

cli::cli_alert_info( "Number of effective outcome traits based on PCA: {effnumb}" )
```

#### PCA plot of variable structure

```{r}
pcaplot = var_pca_fit$x |> as_tibble() |> mutate(k = as.factor(k) ) |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = k), size = 4, shape = 21, alpha = 0.8) +
  geom_text(aes(label = rownames(var_pca_fit$x), color = k), vjust = -1.2, size = 2) +
  theme_minimal() +
  labs(title = "PCA Plot of Variable Structure",
       x = paste0("Principal Component 1 ", " (", round(varexp[1]*100,1), "% Variance Explained)"),
       y = paste0("Principal Component 2 ", " (", round(varexp[2]*100,1), "% Variance Explained)"),
       fill = "Cluster",
       subtitle = "   - clusters defined by tree cut above") + 
  guides(color = "none")
  
## save pca plot to file
f = "../results/figures/variable_correlation_pca_plot.pdf"
pdf(f, width = 8, height = 6)
print(pcaplot)
dev.off()
```


```{r}
## print to screen
pcaplot
```


#### PCA biplot  of individual structure

```{r}
varexp = summary(ind_pca_fit)[[6]][2,]

## 1. Scores (for individuals)
scores <- ind_pca_fit$x[, 1:2] |>
  as_tibble(rownames = "Sample") |>
  select(1:3) 

## k-means on individuals given PC1 and PC2
ind_k = kmeans( scores[, 2:3], centers = 4)$cluster
scores <- scores |> mutate( k = as.factor(ind_k) )

## 2. Loadings (variables)
loadings <- ind_pca_fit$rotation[, 1:2] |>
  as_tibble(rownames = "Variable")

## 3. Select top 5 contributors for each PC (by absolute loading)
top_pc1 <- loadings |>
  arrange(desc(abs(PC1))) |>
  slice_head(n = 5)

top_pc2 <- loadings |>
  arrange(desc(abs(PC2))) |>
  slice_head(n = 5)

# combine, allowing overlap but dropping duplicate rows
top_loadings <- bind_rows(top_pc1, top_pc2) |>
  distinct(Variable, .keep_all = TRUE)

## 4. Scale arrows to fit nicely in score space
arrow_mult <- min(
  (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
  (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
)

top_loadings_scaled <- top_loadings |>
  mutate(
    PC1 = PC1 * arrow_mult * 0.8,
    PC2 = PC2 * arrow_mult * 0.8
  )

## 5. Plot: scores + only top loading arrows
biplot = ggplot() +
  # points (individuals)
  geom_point(
    data = scores,
    aes(x = PC1, y = PC2, fill = k),
    size = 4,
    alpha = 0.8,
    shape = 21
  ) +
  scale_fill_brewer(palette = "Set1") +
  # loading arrows for top features
  geom_segment(
    data = top_loadings_scaled,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.02, "npc")),
    color = "dodgerblue"
  ) +
  # labels for top features
  geom_text(
    data = top_loadings_scaled,
    aes(x = PC1, y = PC2, label = Variable),
    hjust = 0.5,
    vjust = -0.4,
    size = 2
  ) +
  theme_minimal() +
  labs(
    title = "PCA Biplot",
    subtitle = "   - Top contributing variables shown",
    x = paste0("PC1 (Variance Explained = ", round(varexp[1] * 100, 1), "%)"),
    y = paste0("PC2 (Variance Explained = ", round(varexp[2] * 100, 1), "%)"),
    fill = "k-means\nCluster"
  )

```


```{r}
biplot
```

## Testing for a condition effect

```{r}
## Rnank Normal Tranformation function
rnt <- function(x) {
  r <- rank(x, na.last = "keep")
  qnorm( (r - 0.5) / sum(!is.na(x)) )
}

## testing for condition effects
cond_effects <- map_dfr(vars_2_test, \(y) {
  form <- as.formula(paste0("`", y, "` ~ Condition + (1|SubjectID)"))
  # form <- as.formula(paste0("scale(`", y, "`) ~ Condition + (1|SubjectID)"))
  # form <- as.formula(paste0("rnt(`", y, "`) ~ Condition + (1|SubjectID)"))
  fit  <- lmer(form, data = mydata)
  
  shap = shapiro.test(residuals(fit))
  sw_wstat = shap$statistic
  sw_pvalue = shap$p.value
  
  broom.mixed::tidy(fit, effects = "fixed") |>
    mutate(outcome = y, .before = 1) |>
    # mutate(shapiro_w = sw_wstat, shapiro_p = sw_pvalue, .before = estimate)
    mutate(shapiro_w = sw_wstat, shapiro_p = sw_pvalue, .after = p.value)
})

## SCALED Data analysis
scaled_cond_effects <- map_dfr(vars_2_test, \(y) {
  # form <- as.formula(paste0("`", y, "` ~ Condition + (1|SubjectID)"))
  form <- as.formula(paste0("scale(`", y, "`) ~ Condition + (1|SubjectID)"))
  # form <- as.formula(paste0("rnt(`", y, "`) ~ Condition + (1|SubjectID)"))
  fit  <- lmer(form, data = mydata)
  
  shap = shapiro.test(residuals(fit))
  sw_wstat = shap$statistic
  sw_pvalue = shap$p.value
  
  broom.mixed::tidy(fit, effects = "fixed") |>
    mutate(outcome = y, .before = 1) |>
    # mutate(shapiro_w = sw_wstat, shapiro_p = sw_pvalue, .before = estimate)
    mutate(shapiro_w = sw_wstat, shapiro_p = sw_pvalue, .after = p.value)
})
## edit names
colnames(scaled_cond_effects)[-c(1:3)] = paste0("scaled_", colnames(scaled_cond_effects)[-c(1:3)])

## Rank Normal Transformed Analysis
rnt_cond_effects <- map_dfr(vars_2_test, \(y) {
  # form <- as.formula(paste0("`", y, "` ~ Condition + (1|SubjectID)"))
  # form <- as.formula(paste0("scale(`", y, "`) ~ Condition + (1|SubjectID)"))
  form <- as.formula(paste0("rnt(`", y, "`) ~ Condition + (1|SubjectID)"))
  fit  <- lmer(form, data = mydata)
  
  shap = shapiro.test(residuals(fit))
  sw_wstat = shap$statistic
  sw_pvalue = shap$p.value
  
  broom.mixed::tidy(fit, effects = "fixed") |>
    mutate(outcome = y, .before = 1) |>
    # mutate(shapiro_w = sw_wstat, shapiro_p = sw_pvalue, .before = estimate)
    mutate(shapiro_w = sw_wstat, shapiro_p = sw_pvalue, .after = p.value)
})
## edit names
colnames(rnt_cond_effects)[-c(1:3)] = paste0("rnt_", colnames(rnt_cond_effects)[-c(1:3)])


## merge all sum stats together
cond_effects = left_join(cond_effects, scaled_cond_effects, by = c("outcome","effect", "term") )
cond_effects = left_join(cond_effects, rnt_cond_effects, by = c("outcome","effect", "term") )

## print table to screen
cond_effects = cond_effects |> 
  filter(term == "ConditionEX") |>
  mutate( bh_padj = p.adjust(p.value, method = "BH"), .before = shapiro_w ) |>
  mutate( meets_study_alpha = ifelse(p.value < 0.05/10, "yes", "no"), .before = shapiro_w ) |>
  arrange(bh_padj) 

## Write sum stats to file
f = "../results/tables/condition_effects_summary_statistics_March_18_2026.csv"
write.csv(cond_effects, f, row.names = FALSE)

## print to screen
cond_effects |> knitr::kable() |> kableExtra::kable_classic(full_width = FALSE)

```


#### Forest plot of scaled condition effects

```{r, fig.height=8, fig.width=6}
fplot = cond_effects |> 
  filter(term == "ConditionEX") |>
  ggplot( aes(x = reorder(outcome, scaled_estimate), y = scaled_estimate) ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.75) +
  geom_errorbar( aes(ymin = scaled_estimate - 1.96*scaled_std.error, 
                     ymax = scaled_estimate + 1.96*scaled_std.error,
                     color = as.factor(meets_study_alpha)),
                 width = 0.2 , size = 0.5) +
  geom_point( aes(fill = as.factor(meets_study_alpha)), size = 2.5 , shape = 21) +
  scale_fill_brewer(palette = "Set1", name = "meets\nstudy\nalpha") +
  scale_color_brewer(palette = "Set1", name = "meets\nstudy\nalpha") +
  coord_flip() +
  labs(
    x = "Outcome",
    y = "scaled beta: SD units of change under exercise vs rest"
  ) +
  theme_minimal()
fplot
```

```{r}
## save forest plot to file
f = "../results/figures/condition_effects_forest_plot.pdf"
pdf(f, width = 6, height = 8)
print(fplot)
dev.off()
```


## Modeling Energy Intake across meals

### extract meal data 

```{r}
meal_data = mydata |>
  select(SubjectID, Condition, contains( c("Breakfast", "Lunch","Snack", "Diner") )) 
```

### pivot meal data long

```{r}
meal_data <- meal_data |>
  pivot_longer(
    cols = -c(SubjectID, Condition),
    names_to = c("Meal", "Nutrient"),
    values_to = "EI",
    names_pattern = "^(.*)_(.*)_\\(kcal\\)$"
  ) |>
  mutate(
    Meal = case_when(
      Meal == "Diner" ~ "Dinner",   # fix typo if you want
      TRUE ~ Meal
    ),
    Meal = tolower(Meal),
    Nutrient = tolower(Nutrient)
  ) |>
  pivot_wider(
    names_from = Nutrient,
    values_from = EI
  )

```

### edit meal data nomenclature

```{r}
## redfine condition
meal_data <- meal_data |>
  mutate(
    Condition = case_when(
      Condition == "CON"     ~ "rest",
      Condition == "EX" ~ "exercise",
      TRUE ~ Condition
    )
  )

```


### descriptive summary statistics

```{r, message=FALSE, warning=FALSE}
## quantile functions
Q2.5  <- \(x) quantile(x, probs = 0.025, type = 7)
Q97.5 <- \(x) quantile(x, probs = 0.975, type = 7)

tbl_all <- meal_data |>
  select(-SubjectID) |>
  # keep `meal` and `condition` separate
  tbl_strata(
    strata = Meal,
    ~ .x |>
      tbl_summary(
        by = Condition,
        statistic = all_continuous() ~ "{median} ({Q2.5}, {Q97.5})"
      ) |>
      add_p() |>
      bold_labels()
  )

## save tbl_all to file
merged_tbl <- tbl_all |> as_gt()
gtsave(merged_tbl, filename = "../results/tables/meal_summary_statistics.html")
gtsave(merged_tbl, filename = "../results/tables/meal_summary_statistics.png")

```

```{r}
merged_tbl  |>
  tab_options(
    table.width = pct(100),
    table.font.size = 10,
    data_row.padding = px(2)
  )
```


#### Save as a csv file

```{r}
df_export <- tbl_all |> as_tibble()
colnames(df_export) = gsub("\\*", "", colnames(df_export) )
colnames(df_export) = gsub("  \nN = 12", "", colnames(df_export) )

df_export$Characteristic = gsub("__", "", df_export$Characteristic)

colnames(df_export) = paste0( c("", rep("breakfast_",3)  , rep("lunch_",3) , rep("snack_",3) , rep("dinner_",3) ),
                              colnames(df_export))

write.csv(df_export, "../results/tables/meal_summary_statistics.csv", row.names = FALSE)

```


### Testing for a meal condition interaction

```{r}
## define the outcome to run
outcomes = colnames(meal_data)[-c(1:3)]

## run the tests
meal_test_results = lapply(outcomes, function(y) {
  test_meal_condition_interaction(meal_data, outcome = y)
})

## set names
names(meal_test_results) = outcomes

```

### Is there evidence of an interaction effect?

```{r}
## extract interaction p-values
lapply(meal_test_results, function(x) x$interaction_pvalue) |> unlist()
```


### meal specific effects

```{r}
## extract meal specific effects and rbind
meal_effects = lapply(meal_test_results, function(x) x$emm_out) |> 
  bind_rows() |> 
  mutate( bh_padj = p.adjust(p.value, method = "BH") )

## write summary stats to file
f = "../results/tables/meal_specific_Estimate_Marginal_Mean_effect_summary_statistics.csv"
write.csv(meal_effects, f, row.names = FALSE)


meal_effects |> 
  select(Meal, outcome, emmean_exercise, emmean_rest, estimate, SE, df, t.ratio, p.value, bh_padj ) |> 
  arrange(p.value) |>
  knitr::kable() |> kableExtra::kable_classic(full_width = FALSE)
```



## Plot meal specific effects

```{r}
meal_effects |> 
  mutate(Meal = factor(Meal, levels = c("breakfast","lunch","snack","dinner"))) |>
  ggplot( aes(x = Meal, y = estimate, color = outcome) ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.75) +
  geom_point( position = position_dodge(width = 0.5), size = 3.5 ) +
  geom_errorbar( aes(ymin = estimate - 1.96*SE, ymax = estimate + 1.96*SE),
                 position = position_dodge(width = 0.5), width = 0.2 , size = 1) +
  labs(
    x = "Meal",
    y = "Estimated Difference in Energy Intake (kcal)",
    color = "Outcome"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```


## Contrasting choice of LMM frameworks to identify an effect of condition on Dinner Fat

### Fit a LMM for Dinner Fat ~ Condition + (1|SubjectID)

```{r}
fit  <- lmer( `Diner_Fat_(kcal)` ~ Condition + (1|SubjectID), data = mydata)

cli::cli_alert_info( "Summary of LMM fit for Dinner Fat ~ Condition + (1|SubjectID)" )
summary(fit)$coef

cat(paste0("\n\n"))

cli::cli_alert_info( "Estimated Marginal Means and Pairwise Contrasts for Condition Effect on Dinner Fat" )
emmeans(fit, "Condition", contr = "pairwise")

## is the p-value the same as above?
# as.data.frame(emmeans(fit, "Condition", contr = "pairwise")$contrast )[,6] ## Yes !!
```

### Fit a LMM with Meal and Condition interaction for Fat intake

```{r}
fit1 = lmer( fat ~ Condition * Meal + (1 | SubjectID), data = meal_data)
summary(fit1)$coef
```

#### extract emmeans and contrasts for Dinner Fat

```{r}
## estimate the marginal means and contrasts for Dinner Fat
emm = emmeans(fit1, ~ Condition | Meal, contr = "pairwise") 

cli::cli_alert_info( "Estimated Marginal Means for Condition Effect on Dinner Fat from Meal LMM" )
emm$emmeans[3:4]
  
cli::cli_alert_info( "Pairwise Contrasts for Condition Effect on Dinner Fat from Meal LMM" )
emm$contrasts[2]

```






## Session info

```{r}
sessionInfo()
```



